A spatio-temporal analysis of the environmental correlates of COVID-19 incidence in Spain

### 2020-04-20 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.tex
Started working on the results and discussion

\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}

@@

\hypertarget{selection-of-variables}{%
\subsection{Selection of Variables}\label{selection-of-variables}}

@@

\hypertarget{methods-the-spatial-sur-model}{%
\section{Methods: the Spatial SUR
Model}\label{methods-the-spatial-sur-model}}

@@

@@

\hypertarget{exploratory-data-analysis}{%
\subsection{Exploratory Data Analysis}\label{exploratory-data-analysis}}

@@

For estimation, we log-transform the dependent and quantitative
independent variables. The only variable that is not transformed is the
categorical variable for transity systems. A log-log transformation is
appropriate to capture non-linear relationships between variables and
provides a straightforward interpretation of the coefficients as
percentage change. Furthermore, we introduce restrictions so that the
coefficients of two of our control variables are constant over time,
namely GDP per capita and percentage of older adults. We do not see an
\emph{a priori} reason to let those two variables vary across equations,
and the correlation analysis in Figure \ref{fig:daily-correlations} also
suggest little temporal variation. In contrast, we let the spatial
autocorrelation parameter, as well as the parameters of the rest of the
independent variables (including the constant) to vary over
time\footnote{We conducted sensitivity analysis letting all parameters vary over time, and while the results are qualitatively similar, the resulting models are less parsimonious. These results are available in the source R markdown document.}.

After estimation, we compare the goodness of fit of the three SUR
models. Figure \ref{fig:goodness-of-fit} shows the equation-level
coefficient of determination \(R^2\), one for each time period/day. As
well, the overall coefficient of determination for the system is
reported for each model \(\text{pooled}-R^2\). The general trend is
identical for the three models, with the goodness-of-fit improving over
time and plateuing around a value of \(R^2\) of 0.6. Model 1
(\emph{lag8}) performs somewhat better in the first few equations/days,
when the goodness-of-fit is relatively poor, and then again in the last
few equations/days. Model 3 (\emph{lag11w}), in contrast, does not
perform well towards the end of the study period. The most balanced
model in terms of equation-level goodness-of-fit appears to be Model 1
(\emph{lag11}), and this impression is further supported with a slightly
higher value of the \(\text{pooled}-R^2\). This analysis is in line with
the incubation period reported by Lauer et al.~(2020), although not the
use of a weighted average. For the remainder of the paper, we will adopt
Model 2 (\emph{lag11}) as our best model. In the following section we
discuss the results of the analysis in more depth.

@@

Table \ref{tab:summary-best-model} presents a summary.

@@

Figure \ref{fig:results-time-1} shows the temporal evolution of the
spatial autocorrelation coefficient (\(\rho\)) and the intercept of the
model.

@@

Figure \ref{fig:beta-controls-time} shows the temporal evolution of the
coefficients for the two control variables that were not fixed over
time, i.e., \(\log(Density)\) and \(Transit\).

@@

\begin{figure}
\centering
\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/beta-environmenta-time-1.pdf}
\caption{\label{fig:beta-environmental-time}Temporal evolution of
coefficient for the environmental variables; dots are the point
estimates and vertical lines are 95\% confidence intervals. In yellow is
the period after the declaration of the state of emergency, and in
orange is the period when only essential activities were allowed.}
\end{figure}

\hypertarget{concluding-remarks}{%
\section{Concluding Remarks}\label{concluding-remarks}}

@@

@@

@@

>>>>

\setcounter{secnumdepth}{5}
% Pandoc toggle for numbering sections (defaults to be off)

@@

\hypertarget{variables}{%
\subsection{Selection of Variables}\label{variables}}

@@

\hypertarget{methods}{%
\section{Methods: the Spatial SUR Model}\label{methods}}

@@

A note regarding the interpretation of the model is in order. It is
well-known that coefficients in linear regression models are partial
derivatives of the dependent variable with respect to the independent
variables, and therefore directly give the marginal effects or rates of
change. Spatially lagged models, however, are no longer linear. The
intuition behind the non-linearity is that the spatial lag expands the
information set to include information from neighbouring regions: in
other words, the value of an explanatory variable in a spatial unit can
have influence in other spatial units via the spatial lag. This makes
interpretation of the coefficients less straightforward but also richer
(Golgher and Voss, 2016). The results of LeSage (2009) for
cross-sectional spatial lag models can be extended to the spatial SUR
framework. Note that, according to Elhorst (2014), the partial
derivatives have the following interpretation: if an explanatory
variable (\(X_k\)) in a particular province changes, not only the
incidence rate in that province changes, also incidence rates in other
provinces change via the contagion effect. Therefore, a change in
\(X_k\) in a particular province has a \emph{direct effect} on that
province, but also an \emph{indirect effect} on neighbouring provinces.
In this way, the \(i\)th diagonal element of the matrix of partial
derivatives represents the direct effect on the \(i\)th province,
whereas the non-diagonal elements of \(i\)th row of the matrix of
partial derivatives represent the indirect effects on other provinces.
In order to obtain a global indicator, the direct effect is calculated
as the mean of the diagonal elements and captures the average change in
incidence ratio caused by the change in \(X_k\). Likewise, a global
indicator of the indirect effects is calculated as the mean of the
non-diagonal elements. The total effect is the sum of direct and
indirect effects. Finally, note that if \(\rho_k = 0\), the indirect
effects are 0 and the direct effects are equal to \(\beta_kt\).

@@

\hypertarget{eda}{%
\subsection{Exploratory Data Analysis}\label{eda}}

@@

For estimation, we log-transform the dependent and quantitative
independent variables. The only variable that is not transformed is the
categorical variable for transity systems. A log-log transformation is
appropriate to capture non-linear relationships between variables and
provides a straightforward interpretation of the coefficients as
percentage change. Furthermore, we introduce restrictions so that the
coefficients of two of our control variables are constant over time,
namely GDP per capita and percentage of older adults. We do not see an
\emph{a priori} reason to let those two variables vary across equations,
and the correlation analysis in Figure \ref{fig:daily-correlations} also
suggest little temporal variation. In contrast, we let the spatial
autocorrelation parameter, as well as the parameters of the rest of the
independent variables (including the constant) to vary over
time\footnote{We conducted sensitivity analysis letting all parameters vary over time, and while the results are qualitatively similar, the resulting models are less parsimonious. These results are available in the source R markdown document.}.
This will be useful to detect whether there are behavioral adaptations
at the population level over the course of the period examined. As an
example of behavioral adaptations, the effect of density might weaken
over time, in the measure that the effects of the lockdown are felt: at
full compliance with the lockdown, with people practicing social
avoidance, density might matter less than other factors.

After estimation, we compare the goodness of fit of the three SUR
models. Figure \ref{fig:goodness-of-fit} shows the equation-level
coefficient of determination \(R^2\), one for each time period/day. As
well, the overall coefficient of determination for the system is
reported for each model \(\text{pooled}-R^2\). The general trend is
identical for the three models, with the goodness-of-fit improving over
time and plateuing around a value of \(R^2\) of 0.6. Model 1
(\emph{lag8}) performs somewhat better in the first few equations/days,
when the goodness-of-fit is relatively poor, and then again in the last
few equations/days. Model 3 (\emph{lag11w}), in contrast, does not
perform well towards the end of the study period. The most balanced
model in terms of equation-level goodness-of-fit appears to be Model 2
(\emph{lag11}), and this impression is further supported by a slightly
higher value of the \(\text{pooled}-R^2\). The analysis using a lagged
moving average of the environmental variables is in line with the
incubation period reported by Lauer et al.~(2020), although the results
do not support the use of a weighted average. For the remainder of the
paper, we will adopt Model 2 (\emph{lag11}) as our best model. In the
following section we discuss the results of the analysis in more depth.

@@

Table \ref{tab:summary-best-model} presents a summary of the results of
Model 2 (\emph{lag11}). Recall that two coefficients were constrained
and are estimated only for the first equation of the system, and thus
are assumed to be constant over time. These are the coefficients
corresponding to GDP per capita and percentage of older adults. Both
coefficients are significant at \(p<=0.05\). The sign of the coefficient
for GDP per capita is positive, which indicates that wealthier regions
tend to have a higher incidence of COVID-19. This is in line with the
idea that the epidemic started earlier in wealthier places due to their
connections to a globalized world. The sign of the coefficient for
percentage of older adults, on the other hand, is negative. As
previously discussed, the level of social contact of older adults even
under normal circumstances tends to be lower than for younger people. As
a consequence, places with larger populations of older adults appear to
have a natural level of social distancing in place. It is important to
note that this does not detract from evidence that older adults are more
vulnerable individually and in institutional settings, where their case
mortality rates are perhaps the highest of all age groups. Instead, this
result indicates that their presence in the community at large tends to
depress transmission of the virus.

Of the two other control variables, the coefficient of population
density is significant at \(p<=0.05\) in 11 equations, and not
significant in 19. The coefficient for transit is significant at
\(p<=0.10\) in 22 equations, and of those, significant at \(p<=0.05\) in
14 equations. The next four variables are environmental factors. The
coefficient for humidity is significant at \(p<=0.10\) in 20 equations,
and of those, significant at \(p<=0.05\) in 19 equations. Of the
environmental variables, temperature is the only variable that has
significant coefficients in every equation at \(p<=0.05\). Finally,
sunshine has significant coefficients at \(p<=0.10\) in 18 equations,
including 15 coefficients that are significant at \(p<=0.05\).

@@

To better understand the results, we proceed to plot the coefficients in
their temporal sequence. We begin with the evolution of of the spatial
autocorrelation coefficient (\(\rho\)) and the intercept of the model,
both of which are shown in Figure \ref{fig:results-time-1}.

@@

Figure \ref{fig:beta-controls-time} shows the temporal evolution of the
coefficients for the two control variables that were not fixed over
time, i.e., \(\log(Density)\) and \(Transit\). In Section
\ref{variables} we had anticipated a positive sign for the coefficient
of this variable. The negative sign in the results is therefore somewhat
surprising. Upon reflection, however, it seems reasonable that the sign
would be negative during the period under analysis.

@@

\begin{figure}
\centering
\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/beta-environmental-time-1.pdf}
\caption{\label{fig:beta-environmental-time}Temporal evolution of
coefficient for the environmental variables; dots are the point
estimates and vertical lines are 95\% confidence intervals. In yellow is
the period after the declaration of the state of emergency, and in
orange is the period when only essential activities were allowed.}
\end{figure}

The effects of the different variables are summarized in Table
\ref{tab:summary-impacts-best-model}.

\caption{\label{tab:summary-impacts-best-model}\label{tab:summary-impacts-best-model}Summary of direct, indirect, and total effects according to best model (lag11: lagged 11-day moving average)}
\centering
\begin{tabular}[t]{lrrrl}
\toprule
Variable & Min & Mean & Max & Note\\
\midrule
\rowcolor{gray!6}  \addlinespace[0.3em]
\multicolumn{5}{l}{\textbf{Direct Effects}}\\
\hspace{1em}log(GDPpc) & 0.681 & 0.681 & 0.681 & Constrained\\
\hspace{1em}log(Older) & -0.809 & -0.809 & -0.809 & Constrained\\
\rowcolor{gray!6}  \hspace{1em}log(Density) & -0.220 & -0.096 & 0.168 & Non-constrained\\
\hspace{1em}Transit & 0.345 & 0.517 & 0.584 & Non-constrained\\
\rowcolor{gray!6}  \hspace{1em}log(Humidity) & -1.575 & -0.545 & -0.031 & Non-constrained\\
\hspace{1em}log(Temperature) & -2.019 & -1.422 & -0.929 & Non-constrained\\
\rowcolor{gray!6}  \hspace{1em}log(Sunshine) & -0.284 & 0.095 & 0.206 & Non-constrained\\
\addlinespace[0.3em]
\multicolumn{5}{l}{\textbf{Indirect Effects}}\\
\hspace{1em}log(GDPpc) & 0.632 & 0.632 & 0.632 & Constrained\\
\rowcolor{gray!6}  \hspace{1em}log(Older) & -0.751 & -0.751 & -0.751 & Constrained\\
\hspace{1em}log(Density) & -0.032 & -0.004 & 0.156 & Non-constrained\\
\rowcolor{gray!6}  \hspace{1em}Transit & 0.008 & 0.088 & 0.338 & Non-constrained\\
\hspace{1em}log(Humidity) & -1.462 & -0.156 & -0.001 & Non-constrained\\
\rowcolor{gray!6}  \hspace{1em}log(Temperature) & -1.629 & -0.273 & -0.014 & Non-constrained\\
\hspace{1em}log(Sunshine) & -0.263 & -0.006 & 0.028 & Non-constrained\\
\rowcolor{gray!6}  \addlinespace[0.3em]
\multicolumn{5}{l}{\textbf{Total Effects}}\\
\hspace{1em}log(GDPpc) & 1.313 & 1.313 & 1.313 & Constrained\\
\hspace{1em}log(Older) & -1.560 & -1.560 & -1.560 & Constrained\\
\rowcolor{gray!6}  \hspace{1em}log(Density) & -0.244 & -0.100 & 0.324 & Non-constrained\\
\hspace{1em}Transit & 0.500 & 0.606 & 0.772 & Non-constrained\\
\rowcolor{gray!6}  \hspace{1em}log(Humidity) & -3.037 & -0.701 & -0.032 & Non-constrained\\
\hspace{1em}log(Temperature) & -3.385 & -1.695 & -0.946 & Non-constrained\\
\rowcolor{gray!6}  \hspace{1em}log(Sunshine) & -0.547 & 0.089 & 0.228 & Non-constrained\\
\bottomrule
\multicolumn{5}{l}{\textit{Note: }}\\
\multicolumn{5}{l}{Non-constrained: coefficient was allowed to vary across equations}\\
\multicolumn{5}{l}{Constrained: coefficient as constant across equations}\\
\end{tabular}
\end{table}

The temporal variation of the effects is shown in Figure
\ref{fig:plot-impacts}.

\begin{figure}
\centering
\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/plot-impacts-1.pdf}
\caption{\label{fig:plot-impacts}Temporal evolution of effects by date;
the impact is percentage change in incidence as a result of a one
percent change in value of variable.}
\end{figure}

\hypertarget{conclusion}{%
\section{Concluding Remarks}\label{conclusion}}

@@

\leavevmode\hypertarget{ref-Elhorst2014spatial}{}%
Elhorst, J.P., 2014. Spatial econometrics: From cross-sectional data to
spatial panels. Springer.

@@

\leavevmode\hypertarget{ref-Golgher2016interpret}{}%
Golgher, A.B., Voss, P.R., 2016. How to interpret the coefficients of
spatial models: Spillovers, direct and indirect effects. Spatial
Demography 4, 175--205.

@@

\leavevmode\hypertarget{ref-LeSage2009introduction}{}%
LeSage, J., Pace, R.K., 2009. Introduction to spatial econometrics.
Chapman; Hall/CRC.
